---
title: "Common Table Expressions in SQL"
author: "Kevin McManus"
output:
  html_document: default
  html_notebook: default
---

Common Table Expressions (CTE) are an SQL construct that allow you to encapsulate and possibly reuse components of a complex query.
They have many uses, in this example, their main purpose is to simplify what would otherwise be an unwieldy query.

We'll use a specific problem to illustrate the approach.  From a music database that records which cusomters bought which music tracks, we want a query that shows for each country the name, the artist and the number of times purchased for the track that was purchased most frequently in the country.

We'll build the query from the ground up.
Each section below illustrates a different sub-query that is necessary to meet the objective.
The sub-queries are then encapsulated into Common Table Expressions to be reused in later sections of the document.

## Some Preliminaries 
### SQLite 
In this tutorial, we'll use the SQLite database application and a sample database that comes with it.
Download the SQLite database browser from http://sqlitebrowser.org/ and install it on your system.
Obtain the sample database from http://www.sqlitetutorial.net/sqlite-sample-database/ and save it locally.
While you're on this site, examine the database diagram of the sample database.

### Connecting this Notebook to the Database
For the purpose of this tutorial, we need to establish a connection from the database to the R environment that produces this notebook.
Doing so will enable us to issue SQL statements and capture their output in this notebook.
The code below creates a db connection variable `con` which identifies which database we're working with.


```{r, warning=FALSE, message = FALSE}
require(RSQLite)
driver <-  dbDriver("SQLite")

# path to the database file
dbpath <- "./chinook.db"

# connect to the db
con = dbConnect(driver,dbpath)

# test query to make sure the connection is good
dbGetQuery(con,"select CustomerID, FirstName, LastName from customers limit 5")
```
## Constructing the Query 
Our objective is to identify the most popular track and its artist within each country.
Looking at the database diagram, the customer table has a country column.
customers pay invoices, invoices have invoice_items and invoice_items identify the individual tracks.
From the track we can identify the album and therefore the album's artist.
So the query we'll write will combine records from the customer, invoice, invoice_items,  album and artist tables through a series of relational joins.  Five tables will participate in this query.

We could write our query as one big select statment with a whole series of joins.
Doing so would result in a query whose SQL code would be difficult to read and maintain.
Alternately, we could create a series of views for each component of the query and join all the views together.
But this approach results in a lot of views littering the database.

Common Table Expressions (CTE) provide a nice middle ground between these two approaches.

###Common Table Expressions 
Common Table Expressions can function somewhat like a view in the context of a query. (CTE have other uses, too, but these are outside the scope of this discussion.) 
Here's an trivial example of using a CTE to identify the customers in Brazil:
```{sql connection=con}
-- cte's are introduced with the 'WITH' clause:
with Brazilians as (
    select c.country, c.firstname, c.lastname
        from customers c
        where c.Country = 'Brazil'
    )
    
--just reference the cte like any other table or view
select * from Brazilians

```

The general syntax for a CTE is:
```{sql connection=con, eval = FALSE}
CTE ::= WITH cte [,cte ...]

cte ::= <ctename> AS ( <selectstatement> )
```

Now that we've seen an example and know the syntax, let's build our query from the bottom up.

###Tracks and Artists 
Tracks are identified by a trackid; the tracks table has a foriegn key to the album table (the column named albumid).
The albums table likewise has a foriegn key that points to the artist table through the column named artistid.
Since our objective requires us to show track name and artist, we'll make a sub-query to map trackid to trackname and artist. The cell below does this.

Notice in the example below, two CTE strung together separated by a comma.
```{sql connection=con}
--need a (sub)-query to produce trackid, trackname, artist
with
 --get the artist from the album
 artistname as (
    select al.albumid, ar.name as artistname
    from albums al left join artists ar on al.artistid = ar.artistid
    )
    
 -- associate the artist with the track    
,trackartist as (
    select tr.trackid, tr.name as trackname, an.artistname
    from tracks tr left join artistname an on tr.albumid = an.albumid
    )
    
select * from trackartist limit 10

```

###Invoices and Tracks  
Customers live in countries; customers pay invoices and invoices indicate which tracks a customer purchased.
Working from the bottom up, we'll first show for any given invoice which tracks are on that invoice.
```{sql connection=con}
with
  -- which tracks are on which invoices
 invoicetracks as (
    select iv.customerid, iv.invoiceid, ii.trackid
    from invoices iv left join invoice_items ii on iv.invoiceid = ii.invoiceid
 )
 
-- demonstrate which tracks are on which invoice
select * from invoicetracks limit 10
```
The output above shows invoice #98 has two tracks on it, #121 has four, etc and that these were invoiced to customer #1

### Customers and Tracks 
Here we will reuse the CTE from the previous example and add some more customer info to it.
We accomplish this by joining the invoices to customers through the `customerid` column that appears in both tables.
```{sql connection=con, print.limit=100}
with
  -- which tracks are on which invoices
 invoicetracks as (
    select iv.customerid, iv.invoiceid, ii.trackid
    from invoices iv left join invoice_items ii on iv.invoiceid = ii.invoiceid
 )
 
 -- show which tracks each customer purchased and info about that customer:
 select cu.country, cu.firstname, cu.lastname, it.trackid
 from customers cu left join invoicetracks it on cu.customerid = it.customerid
 where cu.country in ('Austria', 'Belgium')
```
The output above shows all of the tracks that were purchased/invoice to customers in Austria and Belgium.
If we then count the country/trackid occurances, we'll be able to see the most popular (commonly purchased) track within each country.

### Track Counts within Country 
Now we have the building blocks to achieve our query's objective.
First we'll look at the number of times *each* track was purchased within each country, then we'll pull out the most popular track for each country.
We'll reuse the `invoicetracks` CTE from the previous query.
```{sql connection=con, max.print = 100}
with
  -- which tracks are on which invoices
 invoicetracks as (
    select iv.customerid, iv.invoiceid, ii.trackid
    from invoices iv left join invoice_items ii on iv.invoiceid = ii.invoiceid
    )
    
 -- the country of  the cusotmer that purchased each track:
,countrytracks as (
    select cu.country, it.trackid
    from customers cu left join invoicetracks it on cu.customerid = it.customerid
    )

--count the number of times each track was purchased within each country
--shown only for Canada and USA for brevity
select ct.country, ct.trackid, count(ct.trackid) as countrytrackcount
    from countrytracks ct
    where country in ('USA', 'Canada')
    group by ct.country, ct.trackid
    order by  countrytrackcount desc
    limit 20
```
Canada has two tracks that were  purchased twice whereas the USA has eight tracks that were purchased twice.  Both countries have a bunch of tracks that were purchased only once.
For Canada, the most popular track is either #1554 or #1563. For the USA, the most popular track is one of the eight shown above.

Our query objective requires us to produce *the* most popular track within a country.
To get the one track that represents the most popular track, we need a way to break the tie between #1554 and #1563 for Canada and the eight tracks for the USA.
Aribtrarily, we'll use the trackid as the tiebreaker. So for Canada, the most popular track is #1563 since its trackid is greater than the other track with two purchases (i.e. #1554).
For the USA, the most popular track is #3455 since it has the greatest trackid among the tracks that have been twice purchased in the USA.


### Get the Max trackid for each country/trackcount pair
We're going to use the `trackid` to break ties in the purchase counts.
Conceptually, we'll compute the number of times each track was purchased within each country, then within each count group (i.e. all the tracks purchased once, twice,..), compute the largest '`trackid`.
Notice in the example below the different `group by` clauses.
```{sql connection=con}
with
  -- which tracks are on which invoices
 invoicetracks as (
    select iv.customerid, iv.invoiceid, ii.trackid
    from invoices iv left join invoice_items ii on iv.invoiceid = ii.invoiceid
    )
    
 -- the country of  the cusotmer that purchased each track:
,countrytracks as (
    select cu.country, it.trackid
    from customers cu left join invoicetracks it on cu.customerid = it.customerid
    )
    
--count the number of times each track was purchased within each country
,countrycounttracks as (
    select ct.country, ct.trackid, count(ct.trackid) as countrytrackcount
    from countrytracks ct
    group by ct.country, ct.trackid
    )

--get the maximum track id within each country & track count group
,countrycountid as (
    select cmt.country, cmt.countrytrackcount, max(cmt.trackid)
    from countrycounttracks cmt
    group by cmt.country, cmt.countrytrackcount
    )
    
-- demonstrate the result:    
select * from countrycountid    
```
Given that we're working with a small sample database, the results of this query aren't particularly interesting since the maximum times a track has been purchased is only one.
Regardless, we press on with our quest!

### Highest Track Count within Country
Notice in the output above, Canada shows two records: one for tracks purchased twice and one for tracks purchased only once.
This table needs to be pared down so that it shows only the maximum `countrytrackcount` for each country.

This is actually more difficult than it may sound because it involves a two-step process.
First, we compute the `countrymaxcount` which is the maximum number of times any track was purchased with a country.
Second, we use the country, maxcount pair to find the trackid (after tie breaking) for that pair.

In the query below, the first step is accomplished by the `countrymaxcount` CTE.
The second step is accomplished by the `countrymaxid` CTE.
Notice the `inner join` in the `countrymaxid` CTE.
This is how we find the `trackid` for the country, maxcount pair.
```{sql connection=con}
with
  -- which tracks are on which invoices
 invoicetracks as (
    select iv.customerid, iv.invoiceid, ii.trackid
    from invoices iv left join invoice_items ii on iv.invoiceid = ii.invoiceid
    )
    
 -- the country of  the cusotmer that purchased each track:
,countrytracks as (
    select cu.country, it.trackid
    from customers cu left join invoicetracks it on cu.customerid = it.customerid
    )
    
--count the number of times each track was purchased within each country
,countrycounttracks as (
    select ct.country, ct.trackid, count(ct.trackid) as countrytrackcount
    from countrytracks ct
    group by ct.country, ct.trackid
    )
    
--for each country, compute the maximum number of times any track was purchased
,countrymaxcount as (
    select ct.country, max(ct.countrytrackcount) as maxcount
    from countrycounttracks ct
    group by ct.country
    )
    
--find the maximum trackid within each country, maxcount pair
,countrycountid as (    
    select cmt.country, cmt.countrytrackcount, max(cmt.trackid) trackid
    from countrycounttracks cmt
    group by cmt.country, cmt.countrytrackcount
    )
    
--get the trackid associated with the country, maxcount pair
,countrymaxid as (
    select cmc.country, cci.trackid, cmc.maxcount
    from countrymaxcount cmc inner join countrycountid cci on
        cmc.country = cci.country and cmc.maxcount = cci.countrytrackcount
    )

--demonstrate
select * from countrymaxid

```


###Put it All Together
All that's left to do now is  get the track name and artist associated with the `trackid`.
To do so, we'll include the `trackartist` and `artistname` CTE that we developed way back in the beginning.
```{sql connection=con, max.print = 100}
with
  -- which tracks are on which invoices
 invoicetracks as (
    select iv.customerid, iv.invoiceid, ii.trackid
    from invoices iv left join invoice_items ii on iv.invoiceid = ii.invoiceid
    )
    
 -- the country of  the cusotmer that purchased each track:
,countrytracks as (
    select cu.country, it.trackid
    from customers cu left join invoicetracks it on cu.customerid = it.customerid
    )
    
--count the number of times each track was purchased within each country
,countrycounttracks as (
    select ct.country, ct.trackid, count(ct.trackid) as countrytrackcount
    from countrytracks ct
    group by ct.country, ct.trackid
    )
    
--find the maximum trackid within each country, maxcount pair
,countrycountid as (    
    select cmt.country, cmt.countrytrackcount, max(cmt.trackid) trackid
    from countrycounttracks cmt
    group by cmt.country, cmt.countrytrackcount
    )
    
--for each country, compute the maximum number of times any track was purchased    
,countrymaxcount as (
    select ct.country, max(ct.countrytrackcount) as maxcount
    from countrycounttracks ct
    group by ct.country
    )
    
--get the trackid associated with the country, maxcount pair
,countrymaxid as (
    select cmc.country, cci.trackid, cmc.maxcount
    from countrymaxcount cmc inner join countrycountid cci on
        cmc.country = cci.country and cmc.maxcount = cci.countrytrackcount
    )
    
 --get the artist from the album
,artistname as (
    select al.albumid, ar.name as artistname
    from albums al left join artists ar on al.artistid = ar.artistid
    )
    
 -- associate the artist with the track    
,trackartist as (
    select tr.trackid, tr.name as trackname, an.artistname
    from tracks tr left join artistname an on tr.albumid = an.albumid
    )

--tada!!    
select cmi.country, cmi.maxcount, cmi.trackid,
        ta.trackname, ta.artistname
from countrymaxid cmi left join trackartist ta on cmi.trackid = ta.trackid
```

### Tidy up 
Like drinking 64 oz of water every day or flossing your teeth after each meal, it is a good practice to close your database connections!
```{r}
# Close the connection
dbDisconnect(con)
```
##Conclusions
Common table expressions allow you to encapsulate the sub-queries of a complex query among their other uses.
By doing so, you can break a query into its logical components and provide a meaningful name for each component.
Building the query from the ground up by accumulating CTE as was done here allows you to verify that each step is providing the results you expect.
The net result is a complex query that easy (easier) to read and maintain.
